Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-08 manuscript/manuscript.Rmd
cleaned up figs

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
devtools::install_github("jhollist/microcystinchla",
                         auth_token = getOption("Github_Access"))
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

@@

################################################################################
# Data Prep
################################################################################
data_url<-"http://water.epa.gov/type/lakes/assessmonitor/lakessurvey/upload/NLA2007_Recreational_ConditionEstimates_20091123.csv"
nla_dat <- get_nla(data_url)
nla_dat <- mutate(nla_dat,log_chla = log(CHLA),log_micro = log(MCYST_TL_UGL))

@@

epa_child_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[1]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_drink_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[2]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

epa_adult_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[3]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_low1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[4]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_low2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[5]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_med1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[6]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_med2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[7]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide arrange of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect agains the impacts, several health advisory levels have been set for some of these toxins, in particular, one of the most common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, measuring micorcystin can not currently be done *in situ* and requires samples be processed in a lab.  This time consuming and expensive.  It is possible to find reliable indicators that may be estimated quickly and *in situ* as a first defense agains high level of microcystin.  In particular, chlorophyll *a* has been shown to be postively associated with microcystin.  In this paper we use this assocation to provide estimates of chlorophyll *a* that if exceeded would be indiciative of a higher likelihood of exceeding select concentrations of microcystin.  Using the 2007 National Lakes Assessment and a conditional probability appoach that has been used in other water quality settings, we idenfify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recerational use.  For microcystin concentrations of 0.3, 1, 1.6, 2. and 4 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, `r mc_chla[4]`, and `r mc_chla[5]`, respectively.  When managing for these various microsystin levels exceed these reproted chlorophyll *a* concentratoins should be a trigger for further testing and possibly managment action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down their municipal water supply due in part to an excess of Microcystin-LR that resulted from a ongoing cyanobacterial harmful algal bloom (cHAB) in Lake Erie.  Since this event, signficant legislation has been passed in the United States and the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR concentrations that would trigger health advisories.  While these levels and association advisories are likey to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on availble measurments of Microcysin-LR which requires taking regular water samples and having those samples process in a lab to determine the toxin concentration [REFS].  This has the potentially to be costly and time consuming both factors which could limit monitoring efforts.  Fortunately, microcystin-LR has been shown to be associated with several other, more easily measured components of water quality.

@@

We evaulated associated chlorohpyll *a* concentrations for an effect for each of the WHO and EPA levels.  These were  0.3, 1, 1.6, 2, 4, 10, and 20 ug/l.

@@

```{r mc_chla_table, results='asis', echo=FALSE}
mc_chla_tbl <- data.frame(Source=mc_names,Microcystin=mc_guides,Chlorophyll=mc_chla)
pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

@@

```{r epa_child_cp_plot,echo=FALSE}
cp_plot(epa_child_cp)
```

@@

```{r who_drink_cp_plot,echo=FALSE}
cp_plot(who_drink_cp)
```

@@

```{r epa_adult_cp ,echo=FALSE}
cp_plot(epa_adult_cp)
```

@@

```{r who_rec_low1_cp, echo=FALSE}
cp_plot(who_rec_low1_cp)
```

@@

```{r who_rec_low2_cp, echo=FALSE}
cp_plot(who_rec_low2_cp)
```
\newpage

```{r who_rec_med1_cp, echo=FALSE}
cp_plot(who_rec_med1_cp)
```
\newpage

```{r who_rec_med2_cp, echo=FALSE}
cp_plot(who_rec_med2_cp)
```
\newpage

>>>>

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
if(!require(microcystinchla)){
  devtools::install_github("jhollist/microcystinchla",
                           auth_token = getOption("Github_Access"))
}
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

@@

################################################################################
# Data Prep
################################################################################
data_url<-"http://water.epa.gov/type/lakes/assessmonitor/lakessurvey/upload/NLA2007_Recreational_ConditionEstimates_20091123.csv"
nla_dat <- get_nla(data_url)
nla_dat <- mutate(nla_dat,log_chla = log10(CHLA),log_micro = log10(MCYST_TL_UGL))

@@

epa_child_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[1]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_drink_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[2]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

epa_adult_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[3]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_low1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[4]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_low2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[5]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_med1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[6]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

who_rec_med2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[7]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

@@

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the most common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, compared to other water quality measures, measurements of microcystin are not common and current field measurement techniques have limited precision and accuracy.  Addressing these issues will take time and resourced.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be postively associated with microcystin.  In this paper we use this assocation to provide estimates of chlorophyll *a* that if exceeded would be indiciative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability appoach that has been used in other water quality settings, we idenfify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recerational use.  For microcystin concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microsystin levels exceeding these reproted chlorophyll *a* concentrations should be a trigger for further testing and possible managment action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down their municipal water supply due in part to an excess of Microcystin-LR that resulted from a ongoing cyanobacterial harmful algal bloom (cHAB) in Lake Erie.  Since this event, signficant legislation has been passed in the United States and the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR concentrations that would trigger health advisories.  While these levels and associated advisories are likey to help mitigate the impacts from harmful algal blooms, they are not without complications.  
One of these complications is that they rely on availble measurments of Microcysin-LR which requires taking regular water samples and having those samples process in a lab to determine the toxin concentration [REFS].  This has the potentially to be costly and time consuming both factors which could limit monitoring efforts.  Fortunately, microcystin-LR has been shown to be associated with several other, more easily measured components of water quality.

@@

We evaulated associated chlorohpyll *a* concentrations for an effect for each of the WHO and EPA levels. Lakes with higher microcystin-LR concentrations were rare. Only `r 100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat))` % of lakes sampled had a concentration greater than 10.  For this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were  0.3, 1, 1.6, and 2 ug/l.

@@

```{r mc_chla_table, results='asis', echo=FALSE}
mc_chla_tbl <- data.frame(Source=mc_names,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

@@

```{r epa_child_cp_plot,echo=FALSE}
plot_cp(epa_child_cp,x = "Log 10 Chlorophyll", y = "Probability of Exceeding EPA Child HA")
```

@@

```{r who_drink_cp_plot,echo=FALSE}
plot_cp(who_drink_cp, x = "Log 10 Chlorophyll", y = "Probability of Exceeding WHO Drinking HA")
```

@@

```{r epa_adult_cp ,echo=FALSE}
plot_cp(epa_adult_cp, x = "Log 10 Chlorophyll", y = "Probability of Exceeding EPA Adult HA")
```

@@

```{r who_rec_low1_cp, echo=FALSE}
plot_cp(who_rec_low1_cp, x = "Log 10 Chlorophyll", y = "Probability of Exceeding WHO Low Probability HA")
```

@@

### 2015-09-08 manuscript/manuscript.tex
cleaned up figs

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
arrange of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect agains the impacts, several health advisory
levels have been set for some of these toxins, in particular, one of the
most common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, measuring
micorcystin can not currently be done \emph{in situ} and requires
samples be processed in a lab. This time consuming and expensive. It is
possible to find reliable indicators that may be estimated quickly and
\emph{in situ} as a first defense agains high level of microcystin. In
particular, chlorophyll \emph{a} has been shown to be postively
associated with microcystin. In this paper we use this assocation to
provide estimates of chlorophyll \emph{a} that if exceeded would be
indiciative of a higher likelihood of exceeding select concentrations of
microcystin. Using the 2007 National Lakes Assessment and a conditional
probability appoach that has been used in other water quality settings,
we idenfify chlorophyll \emph{a} concentrations that are more likely
than not to be associated with an exceedance of a microcystin health
advisory level. We look at the recent US EPA standards for drinking
water as well as the World Health Organization levels for drinking water
and recerational use. For microcystin concentrations of 0.3, 1, 1.6, 2.
and 4 we find chlorophyll \emph{a} concentrations of 23.68, 65.2, 79.8,
113.14, and 273.6, respectively. When managing for these various
microsystin levels exceed these reproted chlorophyll \emph{a}
concentratoins should be a trigger for further testing and possibly
managment action.

@@

We evaulated associated chlorohpyll \emph{a} concentrations for an
effect for each of the WHO and EPA levels. These were 0.3, 1, 1.6, 2, 4,
10, and 20 ug/l.

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.13\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Microcystin
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Chlorophyll
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Child
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
0.3
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
23.68
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
65.2
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Adult
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1.6
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
79.8
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
2
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
113.1
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
4
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
273.6
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
10
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
338.4
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
20
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
338.4
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\includegraphics{manuscript_files/figure-latex/who_rec_low2_cp-1.pdf}
\newpage

\includegraphics{manuscript_files/figure-latex/who_rec_med1_cp-1.pdf}
\newpage

\includegraphics{manuscript_files/figure-latex/who_rec_med2_cp-1.pdf}
\newpage

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
most common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resourced. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be postively associated with
microcystin. In this paper we use this assocation to provide estimates
of chlorophyll \emph{a} that if exceeded would be indiciative of a
higher probability of exceeding select health advisory concentrations
for microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability appoach that has been used in other water
quality settings, we idenfify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recerational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 25.1, 69.6, 84.96, and 113.14, respectively. When
managing for these various microsystin levels exceeding these reproted
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible managment action.

@@

We evaulated associated chlorohpyll \emph{a} concentrations for an
effect for each of the WHO and EPA levels. Lakes with higher
microcystin-LR concentrations were rare. Only 1.1616651 \% of lakes
sampled had a concentration greater than 10. For this analysis we focus
on the microcystin concentrations that are better represented in the NLA
data. These were 0.3, 1, 1.6, and 2 ug/l.

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.13\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Microcystin
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Chlorophyll
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Child
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
0.3
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
25.1
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
69.6
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Adult
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1.6
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
84.96
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
2
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
113.1
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@
